Boosting Higgs Pair Production in the $b\bar{b}b\bar{b}$ Channel at the LHC with Multivariate Techniques

### 2015-07-13 paper/sec-results.tex
Work on paper

\section{Results for the cut-based analysis}

@@

In this section we discuss the results of the traditional
cut-based analysis, using the analysis strategy presented in the previous
section.
%
We compare our results with other recent related studies,
and in particular we study how the signal significance
is modified if only the $4b$ QCD background is taken into account,
but the $2b2j$ and $jjjj$ backgrounds are neglected.
%
Then in the next section we show how using multivariate techniques
one can substantially improve the signal significance as compared
to the cut-based analysis.

\subsection{Signal significance}

First of all we compare the cross-sections at various
analysis levels for the three categories, for signal and background,
by providing a detailed cut-flow.
%
Here we consider all backgrounds enumerated in Sect.~\ref{mcgeneration},
and below we discuss how results are modified if only the $4b$
multijet background is used.
%
In particular,
we consider the following levels of the cut-flow
\begin{itemize}
\item {\bf C0}: cross-sections at the generator
  level, before any analysis cuts.
\item {\bf C1}:  cross-sections after the basic jet acceptance selection
  cuts, but before $b$-tagging.
\item {\bf C2}: cross-sections after after the $b$-tagging.
\end{itemize}
Note that {\bf C1} already includes the effect of the Higgs mass
window cut, Eq.~(\ref{higgsmasswindow}).
%
Of course at the {\bf C0} level the numbers will be the same
for the three individual categories.

In the following, results are presented for the exclusive
categorisation, as discussed in Sect.~\ref{sec:categorisation}.
%
In Sect.~\ref{sec:overlap} we comment
interplay of the overlap between the various categories.
%
At each stage of the cut flow, we also provide the number
of events that would be expected at the HL-LHC with
an integrated luminosity of $\mathcal{L}=3000$ fb$^{-1}$,
as well as 
the signal significance $S/\sqrt{B}$ and the signal
over background ratio $S/B$.
%
We emphasize that it is important not only to achieve a good signal
significance, $S/\sqrt{B}$, but also a good signal over background ratio $S/B$:
if the latter is too small, a measurement of the background with an infeasible small
systematic uncertainty would be required.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}[t]
  \centering
  \begin{tabular}{c|c|c||c|c||c|c}
    \hline
    \multicolumn{7}{c}{Boosted category}\\
    \hline
    \hline
    &    \multicolumn{2}{c||}{$\sigma$ (fb)}   &  \multicolumn{2}{c||}{$N_{\rm ev}$}
    &   $S/\sqrt{B}$  & $S/B$\\
      &    Signal & Back   &  Signal  & Back
    &   & \\
    \hline
        {\bf C0}  &  36.9  & $9.9\,10^{9}$ & $1.1\,10^5$ & $3.0\,10^{13}$  &  0.020 & $3.7\,10^{-9}$\\
        {\bf C1}  &  0.46    & $1.3\,10^6$    &  $1.4\,10^3$   & $3.9\,10^9$     & 0.022     &  $3.5\,10^{-7}$ \\
        {\bf C2}  &  0.064     &  12.6     &  193   &  $3.8\,10^4$    &  0.99    &  $5.1\,10^{-3}$ \\
        \hline
  \end{tabular}
  $\,$\\
  \vspace{0.4cm}
  \begin{tabular}{c|c|c|c|c|c|c}
    \hline
    \multicolumn{7}{c}{Intermediate category}\\
    \hline
    \hline
    &    \multicolumn{2}{c|}{$\sigma$ (fb)}   &  \multicolumn{2}{c|}{$N_{\rm ev}$}
    &   $S/\sqrt{B}$  & $S/B$\\
      &    Signal & Back   &  Signal  & Back
    &   & \\
    \hline
         {\bf C0}  &  36.9  & $9.9\,10^{9}$ & $1.1\,10^5$ & $3.0\,10^{13}$  &  0.020 & $3.7\,10^{-9}$\\
        {\bf C1}  &   1.44    &   $1.1\,10^{7}$  &  $4.3\,10^3$   &  $3.3\,10^{10}$    &  0.024    &  $1.3\,10^{-7}$ \\
        {\bf C2}  &   0.22    &  722   &  660   &   $2.2\,10^6$   &   0.45   &  $3.1\,10^{-4}$ \\
        \hline
  \end{tabular}
  $\,$\\
  \vspace{0.4cm}
  \noindent
  \begin{tabular}{c|c|c|c|c|c|c}
    \hline
    \multicolumn{7}{c}{Resolved category}\\
    \hline
    \hline
    &    \multicolumn{2}{c|}{$\sigma$ (fb)}   &  \multicolumn{2}{c|}{$N_{\rm ev}$}
    &   $S/\sqrt{B}$  & $S/B$\\
      &    Signal & Back   &  Signal  & Back
    &   & \\
    \hline
       {\bf C0}  &  36.9  & $9.9\,10^{9}$ & $1.1\,10^5$ & $3.0\,10^{13}$  &  0.020 & $3.7\,10^{-9}$\\
        {\bf C1}  &   3.34    & $7.5\,10^{7}$    & $1.1\,10^4$    & $3.3\,10^{10}$     & 0.021     & $4.5\,10^{-8}$  \\
        {\bf C2}  &   0.54    &  $3.5\,10^{3}$   &  $1.6\,10^{3}$   &   $1.1\,10^{7}$   & 0.50     &  $1.5\,10^{-4}$ \\
        \hline
  \end{tabular}
  \caption{\small Cut-flow for the analysis of the boosted (top),
    intermediate (middle) and resolved (bottom)
    categories.
    %
    At each level of the cut-flow, we indicate the cross-sections and the number of
    expected events at the HL-LHC for $\mathcal{L}_{\rm int}=3$ ab$^{-1}$, both for
    signal events and for all the backgrounds combined.
    %
    We also provide in each case the
    signal significance $S/\sqrt{B}$ and the signal
    over background ratio $S/B$.
    %
    The first row {\bf C0} is the generator-level result and thus is common
    to all three categories.
    %
    See text for more details.
    \label{table:cutflow}
  }
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

The results for the three categories have been collected in
Table~\ref{table:cutflow}.
%
Let us begin with the discussion of the boosted category, that as we have explained
is the one that benefits from a higher signal significance.
%
We see that after all the cuts, including $b$-tagging,
we end up with almost 200 signal events at the HL-LHC, and
still a substantial background of around $40k$ events.
%
The signal significance is around 1.0 at the end of the
cut-based analysis.
%
While it should have been possible to increase this significance
but using more aggressive cuts, we have refrained doing so
since we prefer this optimisation to per performed by
the MVA.

Note that from the results of Table~\ref{table:cutflow}
it is easy to compute the corresponding numbers
for the end of Run II at the LHC with
$\mathcal{L}_{\rm int}=300$ fb$^{-1}$: in the boosted category,
we have only
19 signal events, and the signal significance drops down to
$S/\sqrt{B}\sim 0.3$.
%
We will discuss in the next section what are the Run II prospects
once we include the effects of the MVA in the analysis, but
it is clear already at this level that, unless rates are
increased by new BSM dynamics, one really needs the full
integrated luminosity of the HL-LHC to be able to observe
the Higgs pair production process.

Both the intermediate and resolved categories benefit from higher signal yields,
specially in the resolved category, but this enhancement is canceled by the stronger
increase in the QCD multi-jet background.
%
We see that in both cases the signal significance is around half of that in the boosted category,
with in addition $S/B$ being an order of magnitude smaller.
%
Therefore, the boosted category is clearly determined to be the most useful category,
specially due to the significant suppression of the QCD multi-jet background.
%
And we still have not exploited all the rich information contained on the jet
substructure, as we will do in the next section.

From the results of Table~\ref{table:cutflow}
we see that after the cut-based analysis, the significance of the Higgs pair production
observation in the $4b$ channel using the boosted topology
is $S/\sqrt{B}=0.99$,
and that the combination
of the boosted, intermediate and resolved categories gives a slightly higher
value, around 1.20.
%
We discuss in the next section how this significance can be enhanced by means
of multivariate analysis.

@@

In summary, we find that the results of our cut-based analysis are roughly
consistent with the results of previous related studies.
%
As we show next, once the MVA is used, we manage to outperform these previous
works.

@@

In Table~\ref{table:cutflowBack}   we show the results for the cross-sections for the different
backgrounds that we consider at the end of the cut-based analysis, for each
of the three categories.
%
All the backgrounds have been normalized to known higher-order results,
as summarized in Table~\ref{tab:samples}.
%
As we can see from this comparison, as opposed that what naively one would expect,
the QCD $2b2j$ background turns out not to be negligible as compared to the $4b$
multijets in any of the categories.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}[t]
  \centering
  \begin{tabular}{c|c|c|c}
    \hline
    Background  &    \multicolumn{3}{c}{$\sigma$ (fb)}   \\
    \hline
  &    Resolved  &  Intermediate  &   Boosted \\
    \hline
    \hline
  QCD $4b$   &   $9.9\,10^{2}$       &  $59$       &   $7.1$        \\
  QCD $2b2j$ &   $2.4\,10^{3}$       &  $6.4\,10^{2}$       &   $3.5$        \\
  QCD $4j$ &     $1.7\,10^{2}$       &   $22$      &    $2.1$        \\
  $t\bar{t}$ &   $3.6$       &  $2.2$       &   $4.5\,10^{-2}$        \\
  \hline
  \hline
Total     &    $3.6\,10^{3}$      &   $7.2\,10^{2}$      &     $13$      \\
\hline
\end{tabular}
    \caption{\small The cross-sections for each of the individual background
      components (and their total sum) at the end of the cut-based
      analysis, for each of three categories.
    \label{table:cutflowBack}
    }
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

In the boosted case, the sum of the $2b2j$ and $4j$ components is almost as large as the
irreducible $4b$ background, thus increasing the total background rates by a factor 2 as compared
to the case where only $4b$ is accounted for.
%
This effect is more dramatic in the intermediate and resolved categories: in the former,
$2b2j$ completely dominates the background (thus the increase in $S/\sqrt{B}$ noted in
Table~\ref{table:cutflow4B}  when only the $4b$ component is kept), and in the latter
it is comparable to $4b$.
%
Therefore, we conclude that an accurate estimate of the fake $b$-tag contribution is
an essential component of this analyses, and that complete QCD multijet samples
should be included.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[t]
\begin{center}
  \includegraphics[width=0.49\textwidth]{plots/pt_H0_C2_boost_back.pdf}
  \includegraphics[width=0.49\textwidth]{plots/pt_H0_C2_res_back.pdf}
  \includegraphics[width=0.49\textwidth]{plots/pt_HH_C2_boost_back.pdf}
  \includegraphics[width=0.49\textwidth]{plots/pt_HH_C2_res_back.pdf}
  \caption{\small
    Upper plots: the decomposition of the QCD multijet background in its various
    components, $4b$, $2b2j$ and $4j$, for the $p_T$ of the leading
    Higgs candidate in the boosted (left plot) and resolved (right plot) categories.
    %
    Lower plots: the same comparison this time for the $p_T^{hh}$ of the
    reconstructed di-Higgs system.
}
\label{fig:histoBack}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%

@@

  \subsection{Overlap between different categories}
\label{sec:overlap}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[t]
\begin{center}
\includegraphics[width=0.65\textwidth]{plots/overlap_categories_C1.pdf}
\caption{\small The fraction of events that satisfy the requirements
  of one or more categories, for signal and background events.
  %
  By construction, there is no overlap between the boosted
  and intermediate categories due to the
   orthogonal jet multiplicity cuts.
}
\label{fig:categorisationHisto}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%

>>>>

\section{Results of the cut-based analysis}

@@

In this section we present the results of the 
cut-based analysis and
compare our results with other recent related studies.
%
We study how the signal significance
is modified if only the $4b$ component of the
QCD multi-jet background is taken into account,
but the $2b2j$ and $4j$ components are neglected.
%
We also study the role of the overlap
between the different categories.

\subsection{Cut flow and signal significance}

First of all we compare the cross-sections at various
analysis levels for the three categories, for signal and background events,
by providing a detailed cut-flow.
%
We consider all backgrounds enumerated in Sect.~\ref{mcgeneration},
then we discuss how results are modified if only the $4b$
component is kept.
%
we consider the following levels of the cut-flow
\begin{itemize}
\item {\bf C0}: cross-sections at the generator
  level, before any analysis cuts.
\item {\bf C1}:  cross-sections after the basic jet acceptance selection
  cuts, but before $b$-tagging.
\item {\bf C2}: cross-sections after after the $b$-tagging.
\end{itemize}
Note that {\bf C1} already includes the effect of the Higgs mass
window cut, Eq.~(\ref{higgsmasswindow}).
%
At the {\bf C0} level the numbers will be the same
for the three individual categories.

In the following, results are presented for the exclusive
categorisation defined in Sect.~\ref{sec:categorisation}.
%
%
At each stage of the cut flow, we also provide the number
of events that would be expected at the HL-LHC with
an integrated luminosity of $\mathcal{L}_{\rm int}
=3000$ fb$^{-1}$,
as well as 
the signal significance $S/\sqrt{B}$ and the signal
over background ratio $S/B$.
%
We emphasize that it is important not only to achieve a good signal
significance, $S/\sqrt{B}$, but also a good signal over background ratio $S/B$:
if the latter is too small, a very accurate
measurement of the background with an infeasible small
systematic uncertainty would be required.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}[t]
  \centering
  \begin{tabular}{c|c|c||c|c||c|c}
    \hline
    \multicolumn{7}{c}{Boosted category}\\
    \hline
    \hline
    &    \multicolumn{2}{c||}{$\sigma$ (fb)}   &  \multicolumn{2}{c||}{$N_{\rm ev}$}
    &   $S/\sqrt{B}$  & $S/B$\\
      &    Signal & Back   &  Signal  & Back
    &   & \\
    \hline
        {\bf C0}  &  36.9  & $9.9\,10^{9}$ & $1.1\,10^5$ & $3.0\,10^{13}$  &  0.020 & $3.7\,10^{-9}$\\
        {\bf C1}  &  0.46    & $1.3\,10^6$    &  $1.4\,10^3$   & $3.9\,10^9$     & 0.022     &  $3.5\,10^{-7}$ \\
        {\bf C2}  &  0.064     &  12.6     &  193   &  $3.8\,10^4$    &  0.99    &  $5.1\,10^{-3}$ \\
        \hline
  \end{tabular}
  $\,$\\
  \vspace{0.4cm}
  \begin{tabular}{c|c|c|c|c|c|c}
    \hline
    \multicolumn{7}{c}{Intermediate category}\\
    \hline
    \hline
    &    \multicolumn{2}{c|}{$\sigma$ (fb)}   &  \multicolumn{2}{c|}{$N_{\rm ev}$}
    &   $S/\sqrt{B}$  & $S/B$\\
      &    Signal & Back   &  Signal  & Back
    &   & \\
    \hline
         {\bf C0}  &  36.9  & $9.9\,10^{9}$ & $1.1\,10^5$ & $3.0\,10^{13}$  &  0.020 & $3.7\,10^{-9}$\\
        {\bf C1}  &   1.44    &   $1.1\,10^{7}$  &  $4.3\,10^3$   &  $3.3\,10^{10}$    &  0.024    &  $1.3\,10^{-7}$ \\
        {\bf C2}  &   0.22    &  722   &  660   &   $2.2\,10^6$   &   0.45   &  $3.1\,10^{-4}$ \\
        \hline
  \end{tabular}
  $\,$\\
  \vspace{0.4cm}
  \noindent
  \begin{tabular}{c|c|c|c|c|c|c}
    \hline
    \multicolumn{7}{c}{Resolved category}\\
    \hline
    \hline
    &    \multicolumn{2}{c|}{$\sigma$ (fb)}   &  \multicolumn{2}{c|}{$N_{\rm ev}$}
    &   $S/\sqrt{B}$  & $S/B$\\
      &    Signal & Back   &  Signal  & Back
    &   & \\
    \hline
       {\bf C0}  &  36.9  & $9.9\,10^{9}$ & $1.1\,10^5$ & $3.0\,10^{13}$  &  0.020 & $3.7\,10^{-9}$\\
        {\bf C1}  &   3.34    & $7.5\,10^{7}$    & $1.1\,10^4$    & $3.3\,10^{10}$     & 0.021     & $4.5\,10^{-8}$  \\
        {\bf C2}  &   0.54    &  $3.5\,10^{3}$   &  $1.6\,10^{3}$   &   $1.1\,10^{7}$   & 0.50     &  $1.5\,10^{-4}$ \\
        \hline
  \end{tabular}
  \caption{\small Cut-flow for the analysis of the boosted (top),
    intermediate (middle) and resolved (bottom)
    categories.
    %
    At each level of the cut-flow, we indicate the cross-sections and the number of
    expected events at the HL-LHC for $\mathcal{L}_{\rm int}=3$ ab$^{-1}$, both for
    signal events and for all the backgrounds combined.
    %
    We also provide in each case the
    signal significance $S/\sqrt{B}$ and the signal
    over background ratio $S/B$.
    %
    The first row {\bf C0} is the generator-level result and thus is common
    to all three categories.
    %
    \label{table:cutflow}
  }
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

The results for the three categories have been collected in
Table~\ref{table:cutflow}.
%
Let us begin with the discussion of the boosted category, the one that
leads to a higher signal significance.
%
We see that after all the cuts, including $b$-tagging,
we end up with almost 200 signal events at the HL-LHC, with
still a substantial background of around $40k$ events.
%
The signal significance is around 1.0 at the end of the
cut-based analysis.
%
While it should have been possible to increase this significance
but using more aggressive cuts, we have refrained doing so
since we prefer this optimisation to be performed at the level of
the MVA.

From the results of Table~\ref{table:cutflow}
it is possible to compute the corresponding numbers
for the end of Run II at the LHC with
$\mathcal{L}_{\rm int}=300$ fb$^{-1}$: in the boosted category,
we have only
19 signal events, and the signal significance drops down to
$S/\sqrt{B}\sim 0.3$.
%
We will discuss in the next section what are the Run II prospects
once we include the effects of the MVA in the analysis, but
it is clear already at this level that one really needs the full
integrated luminosity of the HL-LHC to be able to observe
the Higgs pair production process, unless  of course production rates are
increased by new BSM dynamics

Both the intermediate and resolved categories benefit from higher signal yields,
specially in the resolved category, but this enhancement is washed out by the stronger
increase in the QCD multi-jet background.
%
In both cases the signal significance is around half of that in the boosted category,
with in addition $S/B$ being an order of magnitude smaller.
%
Therefore, the boosted category is clearly determined to be the most useful category,
specially due to the significant suppression of the QCD multi-jet background.
%
And we still have not exploited all the rich information contained on the jet
substructure, as we will do in the next section.

From the results of Table~\ref{table:cutflow}
we see that after the cut-based analysis, the significance of the Higgs pair production
observation in the $4b$ channel using the boosted topology
is $S/\sqrt{B}=0.99$,
and that the combination
of the boosted, intermediate and resolved categories gives a slightly higher
value, around 1.20.
%

@@

In summary, we find that the results of our cut-based analysis are roughly
consistent with the results of previous related studies.
%
Indeed, combining the signal significance in the three categories, we end
up with a final $S/\sqrt{B}\simeq 2.3$.
%
As we show next, once the MVA is used, we achieve a substantial
improvement
over previous works, even after taking into account our rather more
complete modelling of the QCD multijet background.

@@

In Table~\ref{table:cutflowBack}   we show the results for the cross-sections for the different
backgrounds that we consider at the end of the cut-based analysis, for each
of the three categories.
%
Recall that all the backgrounds have been normalized to known higher-order results,
as summarized in Table~\ref{tab:samples}.
%
As we can see from this comparison, as opposed that what naively one would expect,
the QCD $2b2j$ background turns out not to be negligible as compared to the $4b$
multijets in any of the categories.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}[h]
  \centering
  \begin{tabular}{c|c|c|c}
    \hline
    Background  &    \multicolumn{3}{c}{$\sigma$ (fb)}   \\
    \hline
  &    Resolved  &  Intermediate  &   Boosted \\
    \hline
    \hline
  QCD $4b$   &   $9.9\,10^{2}$       &  $59$       &   $7.1$        \\
  QCD $2b2j$ &   $2.4\,10^{3}$       &  $6.4\,10^{2}$       &   $3.5$        \\
  QCD $4j$ &     $1.7\,10^{2}$       &   $22$      &    $2.1$        \\
  $t\bar{t}$ &   $3.6$       &  $2.2$       &   $4.5\,10^{-2}$        \\
  \hline
  \hline
Total     &    $3.6\,10^{3}$      &   $7.2\,10^{2}$      &     $13$      \\
\hline
\end{tabular}
    \caption{\small The cross-sections for each of the individual background
      components (and their total sum) at the end of the cut-based
      analysis, for each of three categories.
    \label{table:cutflowBack}
    }
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

In the boosted case, the sum of the $2b2j$ and $4j$ components is almost as large as the
irreducible $4b$ background, thus increasing the total background rates by a factor 2 as compared
to the case where only $4b$ is accounted for.
%
This effect is more dramatic in the intermediate and resolved categories: in the former,
$2b2j$ completely dominates the background (thus the increase in $S/\sqrt{B}$ noted in
Table~\ref{table:cutflow4B}  when only the $4b$ component is kept), and in the latter
it is comparable to $4b$.
%
Therefore, we conclude that an accurate estimate of the fake $b$-tag contribution is
an essential component of this analyses, and that complete QCD multijet samples
should be included.
%
On the other hand, we find that the hadronic $t\bar{t}$ component of the backround
is smaller than the QCD multijet one.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[t]
\begin{center}
 \includegraphics[width=0.49\textwidth]{plots/pt_H0_C2_res_back.pdf}
 \includegraphics[width=0.49\textwidth]{plots/pt_H0_C2_boost_back.pdf}
   \includegraphics[width=0.49\textwidth]{plots/pt_HH_C2_res_back.pdf}
  \includegraphics[width=0.49\textwidth]{plots/pt_HH_C2_boost_back.pdf}
  \caption{\small
    Upper plots: the decomposition of the QCD multijet background in its various
    components, $4b$, $2b2j$ and $4j$, for the $p_T$ of the leading
    Higgs candidate in the boosted (left plot) and resolved (right plot) categories.
    %
    Lower plots: the same comparison this time for the $p_T^{hh}$ of the
    reconstructed di-Higgs system.
}
\label{fig:histoBack}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%

@@

  {\bf compare parton level with hadron level distributions}

  \subsection{The role of the overlap between different categories}
\label{sec:overlap}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[t]
\begin{center}
\includegraphics[width=0.65\textwidth]{plots/overlap_categories_C1.pdf}
\caption{\small The fraction of events that satisfy the requirements
  of one or more categories, for signal and background events.
  %
  By construction, there is no overlap between the boosted
  and intermediate categories due to the
   orthogonal jet multiplicity cuts. {\bf update including event weights}
}
\label{fig:categorisationHisto}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%
